### 2022-09-25 AISTATS2023/paper.tex
updated method description

\newcommand{\dfdx}{\frac{\partial f}{\partial x_s}}
\newcommand{\xc}{\mathbf{x_c}}
\newcommand{\DY}{\mathbf{\Delta Y}}
\newcommand{\xb}{\mathbf{x}}
\newcommand{\Xcb}{\mathbf{X}_c}
\newcommand{\Xb}{\mathcal{X}}

@@

Recently, ML has flourished in critical domains, such as healthcare
and finance. In these areas, we need ML system with the capability to
explain their predictions, apart from predicting with accuracy. For
this reason there is an increased interest in Explainable AI (XAI),
the field that provides interpretations about the behavior of complex
black-box models. XAI literature distinguishes between local and
global explainability
techniques~\citep{Molnar2020interpretable}. Local methods explain a
specific prediction, whereas global methods explain the entire model
behavior. Global methods provide a universal explanation, summarizing
the numerous local ones into a single interpretable outcome, usually a
number or a plot. If a user wants to get a rough overview about which
features are significant (feature importance) or whether a particular
feature has a positive or negative effect on the output (feature
effect), they should opt for a global explainability technique. On
the other hand, aggregating the individual explanations for producing
a concise global one is vulnerable to misinterpretations; Under strong
feature interactions, the global explanation may obfuscate
heterogeneous effects~\citep{herbinger2022repid} that exist under the
hood; a phenomenon called aggregation bias~\citep{mehrabi2021survey}.

@@

However, ALE faces two crucial drawbacks. First, it does not inform
the user about potential heterogeneous effects hidden behind the
average effect. In contrast, in the case of PDP, the heterogeneous
effects can be spotted by exploring the Individual Conditional
Expectations (ICE)\citep{goldstein2015peeking}. Second, ALE
approximation from the samples of the training set requires
partitioning the axis of the feature of interest in a sequence of
non-overlapping intervals and estimating a single effect from the
population of samples in each interval. We refer to this step using
the term \textit{bin-splitting problem}. Deciding on the sequence of
intervals has important implications because ALE's interpretation in
cases of correlated features is meaningful only inside each interval
\citep{molnar2022}. However, this crucial step has not been given the
appropriate attention, making the approximation vulnerable to
potential misinterpretations.

In this paper, we extend ALE with a probabilistic component for
measuring the uncertainty of the explanation. The uncertainty of the
global explanation expresses how certain we are that the averaged
explanation is valid if applied to an instance drawn at random and
reveals the heterogeneous effects. Given that ALE's interpretation
(expected effect and uncertainty) is meaningful only inside each
interval, we design a principled framework, where we treat the
bin-splitting step as a data-driven clustering problem, searching for
the optimal splitting given available instances of the training
set. We, also, present a computationally-grounded algorithm for
finding the optimal solution.

@@

\paragraph{Notation.} We refer to random variables (rv) using
uppercase \( X \), whereas to simple variables with plain lowercase
\( x \). Bold denotes a vector; \( \xb \) for simple variables or
\(\mathbf{X}\) for rvs. Often, we partition the input vector
\(\xb \in \mathbb{R}^D\) to the feature of interest
\(x_s \in \mathbb{R} \) and the rest of the features
\(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
\((x_s, \mathbf{x}_c)\), but we clarify that it corresponds to the
vector \((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote
the corresponding rv as \(X = (X_s, \mathbf{X}_c)\). The black-box
function is \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the
FE of the \(s\)-th feature is
\(f^{\mathtt{<method>}}(x_s)\), where \(\mathtt{<method>}\) is the
name of the FE method.\footnote{An extensive list of all
  symbols used in the paper is provided in the helping material.}

\paragraph{Feature Effect Methods.} The three well-known feature
effect methods are: PDP, MP and ALE. PDP formulates the FE
of the \(s\)-th attribute as an expectation over the marginal
distribution \(\mathbf{X}_c\), i.e.,
\(f^{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\), whereas MP
formulates it as an expectation over the conditional
\(\mathbf{X}_c|X_s\), i.e.,
\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|X_s = x_s}[f(x_s,
\mathbf{X}_c)]\). ALE computes the global effect at \(x_s\) as the
accumulation of the averaged local effects:
\begin{equation}
  \label{eq:ALE_accumulated_mean}
  f^{\mathtt{ALE}}(x_s) = \int_{z_{s,min}}^{x_s} \mathbb{E}_{\Xcb|X_s=z}\left[\frac{\partial f(z, \Xcb)}{\partial z}\right] \partial z
\end{equation}
%
ALE has specific advantages which gain particular value in cases of
correlated input features. In these cases, PDP integrates over
unrealistic instances, due to the use of the marginal distribution
\(\mathbf{X}_c \), and MP computes aggregated effects, i.e., imputes
the combined effect of sets of features to a single feature. ALE
manages to resolve both issues, and is therefore a trust able FE method
in cases of correlated features.

\paragraph{Quantify the Heterogeneous Effects.}

FE methods answer the question \textit{what is expected to happen to
  the output (expected effect), if the value of a specific feature is
  increased/decreased}. Having an answer to the question above, it
comes naturally to also wonder \textit{how certain we are about the
  expected change (uncertainty)}. For this reason, the quantification
of the uncertainty along with the expected effect has attracted a lot
of interest. The level of uncertainty is mostly quantified by
measuring the existence of heterogeneous effects, i.e. whether there
are local explanations that deviate from the expected global
effect. ICE and d-ICE plots provide a visual understanding of the
heterogeneous effects on top of PDPs. Another approach targets on
grouping the heterogeneous effects, e.g., allocating ICE plots in
homogeneous clusters, by dividing the input
space.\citep{molnar2020model} Some other approaches, like H-Statistic,
Greenwel, move a step behind and try to quantify the level of
interaction between the input features, a possible cause of
heterogeneous effects. In this case, the interpretation is indirect,
since a strong interaction index is only an indicator of heterogeneous
effects. The aforementioned approaches are under two limitations; They
either do not quantify the uncertainty of the FE directly or they are
based on PDPs, and, therefore, they are subject to the failure modes
of PDPs in cases of correlated features. To the best of our knowledge,
no method so far targets on quantifying the uncertainty of ALE.

\paragraph{Bin-Splitting for ALE estimation.}

In real ML scenarios, the FE is estimated from the limited instances
of the training set.  \citep{apley2020visualizing} proposed dividing
the \(s\)-th axis in \(K\) bins and estimating the local effects in
each bin by evaluating the black box-function at the bin limits:
\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}(x_s) = \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \left [ f(z_{k}, \xc^i) - f(z_{k-1}, \xc^i)) \right ]
\end{equation}
We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
the set of training instance that lie in the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
\). Afterwards, (cite) proposed the Differential ALE (DALE) for
computing the local effects on the training instances using auto-differentiation:
\begin{equation}
  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \frac{\partial f}{\partial x_s}(\mathbf{x}^i)
\end{equation}
%
Their method has the advantages of remaining on-distribution even when
bins become wider and, most importantly, allows the recomputation of
the accumulated effect with different bin-splitting with near-zero
computational cost. However, none of the approximations above deals
with the crucial problem of bin-splitting. As indicated
by~\citep{molnar2022}, in ALE the effects are computed per interval
(region) and the interpretation of the effect can only be
local.

@@

The NAME method extends ALE for measuring the uncertainty of the
expected effect and automatically extracts regions with homogeneous
effects. In Section~\ref{sec:NAME-definition} we define the component
for the uncertainty quantification. In
Section~\ref{sec:interval-based-estimation}, we show how to estimate
the average effect and the uncertainty on a specific interval from the
limited samples of the training set. We, also, make an important proof
about the aggregated variance defined over a bin. In
Section~\ref{sec:interval-spliting}, we define and solve the problem
of optimal bin-splitting for automatically extract regions with
homogeneous effects. Finally, in Section~\ref{sec:visualization}, we
illustrate the appropriate visualization of NAME for facilitating its
interpretation by a non-expert and we discuss important aspects of the
method.

\subsection{Uncertainty Quantification}
\label{sec:NAME-definition}

ALE defines the local effect of the \(s\)-th feature at point
\((x_s, \xc)\) as \(\dfdx (x_s, \xc)\). All the local explanations at
\(x_s\) are, then, weighted by the conditional distribution
\(p(\xc|x_s)\) and are averaged, to produce the summarized effect at
\(x_s\):
\begin{equation}
  \label{eq:ALE_mean}
  \mu(x_s) = \mathbb{E}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb)\right ]
\end{equation}
%
The FE at \(x_s\) is the accumulation of the averaged local effects
from \(x_{s, min}\)\footnote{\(x_{s,min}\) is the minimum value the
  \(s\)-th attribute can take. In estimations, it is the minimum
  \(s\)-th attribute value among the training set instances.} until
\(x_s\), i.e.
\(f^{\mathtt{ALE}}(x_s) = \int_{x_{s, min}}^{x_s} \mu(z) \partial z
\). Limiting the explanation to the expected value level does not
shed light to potential heterogeneous effects behind the averaged
explanation. Therefore, we model the uncertainty of the local effects
\(\mathcal{H}(x_s)\) with the variance of the local explanations:
\begin{equation}
  \label{eq:ALE_var}
  \mathcal{H}(x_s) := \sigma^2(x_s) = \mathrm{Var}_{\Xcb|x_s}\left [\dfdx (x_s, \Xcb) \right ]
\end{equation}
\noindent
The uncertainty of the explanation emerges from the natural
characteristics of the experiment, i.e.,~the data generating
distribution and the properties of the black-box function. In
Section~\ref{sec:visualization}, we propose appropriate visualizations
for easier interpretation of Eq.~(\ref{eq:ALE_var}). ALE defines the
FE at \(x_s\) as the accumulation of the averaged local effects from
\(x_{min}\) until \(x_s\). Equivalently, we define the accumulated
uncertainty (variance) until \(x_s\), as the integral of the variances
of the local effects:
\begin{equation}
  \label{eq:ALE_accumulated_var}
  f^{\mathtt{ALE}}_{\sigma^2}(x_s) = \int_{z_{s, min}}^{x_s} \sigma^2(z) \partial z
\end{equation}
\noindent
The accumulated uncertainty is not a directly interpretable quantity,
since it measures the total uncertainty that characterizes the
\(s\)-th feature effect plot. As stated in the Introduction, ALE
outcomes should be interpreted at a local level, therefore, the only
\(\mathcal{H}(x_s)\) is the directly interpretable quantity. However,
the accumulated uncertainty will help us define a sensible objective
for the bin-splitting step, as we will discuss in
Section~\ref{sec:interval-spliting}.

\subsection{Interval-Based Estimation}
\label{sec:interval-based-estimation}

In real scenarios, we have ignorance about the data-generating
distribution \(p(x_s, \mathbf{x}_c)\), so, the estimations are based
on the limited instances of the training set. Estimating
Eqs.~\eqref{eq:ALE_mean},~\eqref{eq:ALE_var} at the granularity of a
point is impossible, because the probability of observing a sample
inside the interval \([x_s - h, x_s + h]\) tends to zero, when
\(h \to 0\). We are, therefore, obliged to split the axis of \(x_s\)
into a sequence of non-overlapping intervals (bins) and estimate the
mean and the variance from the samples that lie inside each bin. The
mean effect at an interval \([z_1, z_2)\) is defined as the mean of
the expected effects:
\begin{equation}
  \label{eq:mu_bin}
  \mu(z_1, z_2) = \frac{1}{z_2 - z_1} \int_{z_1}^{z_2}
  \mathbb{E}_{\xc|x_s=z}\left [\frac{\partial f}{\partial x_s} \right ] \partial z
\end{equation}
\noindent
Similarly, the accumulated variance at an interval \([z_1, z_2)\) is
defined as:
\begin{equation}
  \label{eq:var_bin}
  \sigma^2(z_1, z_2) = \int_{z_1}^{z_2}
  \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial
      f}{\partial x_s} - \mu(z_1, z_2) )^2 \right] \partial z
\end{equation}
\paragraph{Theorem 1.} If we define the residual \(\rho(z)\) as the
difference between the expected effect at \(x_s\) and the mean
expected effect at the interval, i.e
\(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the accumulated variance
\(\sigma^2(z_1, z_2)\) equals to the accumulation of the point-wise
variances plus the squared residuals inside the interval:
\begin{equation}
 \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z
\end{equation}
%
The proof is in the Appendix. Theorem 1 decouples the total
uncertainty inside a bin
\(\mathcal{H}_{bin}(z_1, z_2) := \sigma^2(z_1, z_2)\) into two terms. The first term quantifies
the accumulated uncertainty due to the natural characteristics of the
experiment, i.e.,
\(\mathcal{H}(z_1, z_2) = \int_{z_1}^{z_2}\mathcal{H}(z) \partial z\),
and the second term adds extra nuisance uncertainty due to limiting
the resolution
\(\mathcal{H}_n(z_1, z_2) := \int_{z_1}^{z_2}\rho^2(z) \partial z
\). In other words, the creation of a bin burdens the estimation with
a nuisance uncertainty:
\begin{equation}
  \label{eq:uncertainty_bin}
  \mathcal{H}_{bin}(z_1, z_2) = \mathcal{H}(z_1, z_2) + \mathcal{H}_n(z_1, z_2)
\end{equation}
Eqs.~\eqref{eq:mu_bin},~\eqref{eq:var_bin} can be directly estimated
from the set \(\mathcal{S}\) of the dataset instances with the
\(s\)-th feature lying inside the interval, i.e.,
\( \mathcal{S}= \{ \mathbf{x}^i : z_1 \leq x^i_s < z_2 \} \). The mean
effect at the interval, Eq.~(\ref{eq:mu_bin}) is approximated by:
\begin{equation}
  \label{eq:mean_estimation}
  \hat{\mu}(z_1, z_2) = \frac{1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}} \left [ \dfdx(\mathbf{x}^i) \right ]
\end{equation}

and the accumulated variance, Eq.~(\ref{eq:var_bin}) can be
approximated by
\begin{equation}
  \label{eq:variance_estimation}
  \hat{\sigma}^2(z_1, z_2) = \frac{z_2 - z_1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}} \left ( \dfdx(\mathbf{x}^i) - \hat{\mu}(z_1, z_2) \right )^2
\end{equation}
The approximation is unbiased only if the points are uniformly
distributed in \([z_1, z_2]\). Elaborate.

@@

As we discussed in the Introduction, when the features are (strongly)
correlated, ALE provides meaningful interpretations only inside a
bin. The accumulation of the mean effects inside each bin is done only
for a producing smooth curve for easier visual exploration. Therefore,
the extraction of regions with common effects is particularly
important for a correct interpretation.

>>>>

\newcommand{\dfdx}{f^s}
\newcommand{\xc}{\mathbf{x}_c}
\newcommand{\DY}{\mathbf{\Delta Y}}
\newcommand{\xb}{\mathbf{x}}
\newcommand{\Xcb}{\mathbf{X}_c}
\newcommand{\Xb}{\mathcal{X}}

@@

Recently, ML has flourished in critical domains, such as healthcare
and finance. In these areas, we need a combination of accurate
predictions along with meaningful explanations to support them. For
this reason there is an increased interest in Explainable AI (XAI),
the field that provides interpretations about the behavior of complex
black-box models. XAI literature distinguishes between local and
global explainability
techniques~\citep{Molnar2020interpretable}. Local methods explain a
specific prediction, whereas global methods explain the entire model
behavior. Global methods provide a universal explanation, summarizing
the various local explanations into a single interpretable outcome,
usually a number or a plot. If a user wants to get a rough overview
about which features are significant (feature importance) or whether a
particular feature has a positive or negative effect on the output
(feature effect), they should opt for a global explainability
technique. On the other hand, aggregating the individual explanations
for producing a concise global one is vulnerable to
misinterpretations. Under strong interactions and correlations between
features, the global explanation may obfuscate heterogeneous effects
that exist under the hood~\citep{herbinger2022repid}; a phenomenon
called aggregation bias~\citep{mehrabi2021survey}.

@@

However, ALE faces two crucial limitations, the first concerns ALE
definition and the second one the ALE approximation. Regarding the
definition, i.e., the formulation of the feature effect, ALE does not
inform the user about the level of heterogeneous effects hidden behind
the average effect (global explanation). In contrast, in the case of
PDP, the heterogeneous effects can be (partially) spotted by exploring
the Individual Conditional Expectations
(ICE)\citep{goldstein2015peeking}. Regarding the approximation,
e.g.~the estimation of ALE from the limited samples of the training
set, ALE requires an important additional step, called
\textit{bin-splitting}. Bin-splitting consists of partitioning the
axis of the feature of interest in a sequence of non-overlapping
intervals and estimating a single effect from the population of
samples in each interval. Specifying an approriate sequence of
intervals is of particular importance, since ALE's interpretation is
meaningful only inside each interval \citep{molnar2022}. However, this
crucial step has not raised the appropriate attention, making the
approximation vulnerable to potential misinterpretations.

In this paper, we extend ALE with a probabilistic component for
measuring the uncertainty of the explanation. The uncertainty of the
global explanation, i.e.~how certain we are that the averaged
explanation is valid if applied to an instance drawn at random,
quantifies the level of heterogeneous effects. Given that ALE's
interpretation (expected effect and uncertainty) is meaningful only
inside each interval, we design a principled framework, where we treat
the bin-splitting step as a data-driven clustering problem, searching
for the optimal splitting given available instances of the training
set. We, also, present a computationally-grounded algorithm for
finding the optimal solution.

@@

At Section~\ref{sec:feat-effect-meth}, we present the basic Feature
Effect (FE) methods and we review the ALE
definition\citep{apley2020visualizing}. At
Section~\ref{sec:quant-heter-effects}, we present the techniques for
quantifying the heterogeneous effects and we discuss some of their
limitations. Finally, at Section~\ref{sec:ale-approximation}, we
review the ALE approximation\citep{apley2020visualizing}(add us),
describing why it is vulnerable to a poor approximation due to
inappropriate bin-splitting.

\paragraph{Notation.} We refer to random variables (rv) using
uppercase \( X \), to simple variables with plain lowercase \( x \)
and to vectors with bold \( \xb \). Often, we partition the input
vector \(\xb \in \mathbb{R}^D\) to the feature of interest
\(x_s \in \mathbb{R} \) and the rest of the features
\(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
\((x_s, \mathbf{x}_c)\), but we clarify that it implies the vector
\((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote the
corresponding rv as \(\mathbf{X} = (X_s, \mathbf{X}_c)\). The
black-box function is \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and
the FE of the \(s\)-th feature is \(f^{\mathtt{<method>}}(x_s)\),
where \(\mathtt{<method>}\) is the name of the FE method.\footnote{An
  extensive list of all symbols used in the paper is provided in the
  helping material.}

\subsection{Feature Effect Methods And ALE Definition.}
\label{sec:feat-effect-meth}

The three well-known feature effect methods are: PDP, MP and ALE. PDP
formulates the FE of the \(s\)-th attribute as an expectation over the
marginal distribution \(\mathbf{X}_c\), i.e.,
\(f^{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\), whereas MP
formulates it as an expectation over the conditional
\(\mathbf{X}_c|X_s\), i.e.,
\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|X_s = x_s}[f(x_s,
\mathbf{X}_c)]\). ALE defines the local effect of the \(s\)-th feature
at point \(x_s = z\) as \(f^s(z, \xc) = \dfdx (z, \xc)\). All the
local explanations at \(z\) are, then, weighted by the conditional
distribution \(p(\xc|x_s=z)\) and are averaged, to produce the
averaged effect \(\mu(z)\). Finally, ALE is the accumulation of the
averaged local effects:

\begin{equation}
  \label{eq:ALE}
  f^{\mathtt{ALE}}(x_s) = \int_{x_{s,min}}^{x_s} \underbrace{\mathbb{E}_{\Xcb|X_s=z}\left [f^s (z, \Xcb)\right ]}_{\mu(z)} \partial z
\end{equation}

ALE has specific advantages which gain particular value in cases of
correlated input features. In these cases, PDP integrates over
unrealistic instances due to the use of the marginal distribution
\(\mathbf{X}_c \), and MP computes aggregated effects, i.e., imputes
the combined effect of sets of features to a single feature. ALE
manages to resolve both issues, and is therefore the most trustable
method in cases of correlated features.

\subsection{Quantification Of Heterogeneous Effects.}
\label{sec:quant-heter-effects}

FE methods pose the question \textit{what is expected to happen to the
  output (expected effect), if the value of a specific feature is
  increased/decreased}. Having quantified the expected effect, it
comes naturally to also ask \textit{how certain we are about the
  expected change (uncertainty)}. For this reason, the quantification
of the uncertainty, whether there are local explanations that deviate
from the expected global effect, has attracted a lot of interest.

ICE and d-ICE\citep{goldstein2015peeking} provide a set of curves that
are produce a plot plot on top-of PDP. Both methods produce one curve
for each instance of the training set, i.e. ICE the instance effect
\(f^{\mathtt{ICE}}_i(x_s) = f(x_s, \xc^i)\) and d-ICE the derivative
effect, i.e.
\(f^{\mathtt{d-ICE}}_i(x_s) = \frac{\partial f}{\partial x_s} (x_s,
\xc^i)\). The user can visually examine the level of homogeneity of
the set of curves as an indicator of the level of heterogeneous
effects. Some methods try to automate the aforementioned visual
exploration, by grouping (d-)ICE into clusters~\citep{molnar2020model,
  herbinger2022repid}. Some other approaches, like H-Statistic,
Greenweal, quantify the level of interaction between the input
features. A strong interaction index is an indirect indicator for the
existence of heterogeneous effects.

The aforementioned approaches are under two limitations; They either
do not quantify the uncertainty of the FE directly or they are based
on PDPs, and, therefore, they are subject to the failure modes of PDPs
in cases of correlated features, as we will analyze later. To the best
of our knowledge, no method so far quantifies the heterogeneous
effects on-top of ALE.

\subsection{ALE Approximation.}
\label{sec:ale-approximation}

In real ML scenarios, the FE is estimated from the limited instances
of the training set. Therefore, \citep{apley2020visualizing} proposed
dividing the \(s\)-th axis in \(K\) bins and estimating the local
effects in each bin by evaluating the black box-function at the bin
limits:
\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}(x_s) = \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \left [ f(z_{k}, \xc^i) - f(z_{k-1}, \xc^i)) \right ]
\end{equation}
We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
the set of training instance that lie in the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
\). Afterwards, (cite) proposed the Differential ALE (DALE)
approximation for computing the local effects on the training
instances using auto-differentiation:
\begin{equation}
  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \frac{\partial f}{\partial x_s}(\mathbf{x}^i)
\end{equation}
%
Their method has the advantages of remaining on-distribution even when
bins become wider and, most importantly, allows the recomputation of
the accumulated effect with different bin-splitting with near-zero
computational cost.

Both approximations ask from the user to blindly decide the number of
bins \(K\) for splitting the axis into K equally-sized bins, an
approach with crucial limitations. Firstly, it is vulnerable to poor
bin-effect approximations that lead to a misleading ALE plot. On the
one hand, setting a \(K\) to a small value may hide fine-grain effects
due to large bins and setting \(K\) to a high value leads to
estimations of bin-effect from very limited samples. In general, the
user may face contradictory explanations for differnt \(K\) without a
clue to decide which one to trust. Secondly, as indicated
by~\citep{molnar2022}, in ALE the interpretation of the effect can
only be local. Preseting a fixed-size intervals, does not permit the
extraction of local regions (intervals) ng of data-driven local
regions

@@

The NAME method extends ALE for quantifying the level of heterogeneous
effects (uncertainty) and automating the bin-splitting step for robust
estimations. At Section~\ref{sec:NAME-definition}, we define NAME,
presenting the component that extends ALE with uncertainty
quantification. In Section~\ref{sec:interval-based-estimation}, we
provide an alternative definition of NAME based on variable-size bins,
providing important remarks and proofs for connecting it with the
initial definition. In Section~\ref{sec:interval-spliting}, we define
the problem of optimal bin-splitting, i.e. how to optimally
approximate METHOD based on limited samples, and we provide an
algorithmic solution to the problem. Finally, in
Section~\ref{sec:visualization}, we illustrate the appropriate
visualization of NAME for facilitating its interpretation by a
non-expert and we discuss important aspects of the method.

\subsection{NAME: ALE With Uncertainty Quantification}
\label{sec:NAME-definition}

ALE does not shed light to potential heterogeneous effects that
contribute to the average explanation. Therefore, we extend ALE
definition of Eq.~(\ref{eq:ALE}) with a component that quantifies the
level of heterogeneous effects. We denote \(\mathcal{H}(z)\) the
uncertainty of the local effects at a specific point \(x_s=z\) and we
quantify it as the standard deviation of the local explanations:
\begin{equation}
  \label{eq:ALE_var}
  \mathcal{H}(z) := \sigma(z) = \sqrt{\mathbb{E}_{\Xcb|z}\left [ \left (\dfdx (z, X_c) - \mu(z) \right )^2 \right ] }
\end{equation}
\noindent
The uncertainty emerges from the natural characteristics of the
experiment, i.e.,~the feature correlations existent in the data
generating distribution and the implicit interactions of the black-box
function. We also define the accumulated uncertainty at \(x_s\), as
the accumulation of the standard deviation of the local effects along
the axis:
\begin{equation}
  \label{eq:ALE_acc_unc}
  f^{\mathtt{ALE}}_{\sigma}(x_s) = \int_{x_{s, min}}^{x_s} \sigma(z) \partial z
\end{equation}
\noindent
NAME method formulates the effect at a specific point \(x_s\) with a
compact entity that consists of the average effect and the uncertainty
\(\mu(z) \pm \sigma(z)\) and visualizes them as a continuous curve
with a confidence region, i.e.
\(f^{\mathtt{METHOD}}(x_s) := f^{\mathtt{ALE}}_{\mu}(x_s) \pm
f^{\mathtt{ALE}}_{\sigma}(x_s)\). In Section~\ref{sec:visualization},
we propose an appropriate visualization for easier interpretation of
NAME.

\subsection{Interval-Based Formulation of NAME}
\label{sec:interval-based-estimation}

In real scenarios, the estimations are based on the limited instances
of the training set. Estimating \(\mu(z), \sigma(z)\) at the
granularity of a point is impossible, because the probability of
observing a sample inside the interval \([x_s - h, x_s + h]\) tends to
zero, when \(h \to 0\). For this reason, we firstly define the mean
effect (bin-effect) and the uncertainty (bin-uncertainty) inside an interval \([z_1, z_2)\):

\begin{equation}
  \label{eq:mu_bin}
  \mu(z_1, z_2) = \frac{1}{z_2 - z_1} \int_{z_1}^{z_2}
  \mathbb{E}_{\xc|z}\left [f^s(z, \xc) \right ] \partial z
\end{equation}

\begin{equation}
  \label{eq:var_bin}
  \sigma(z_1, z_2) = \frac{\int_{z_1}^{z_2}
  \mathbb{E}_{\xc|x_s=z} \left [ (\frac{\partial
      f}{\partial x_s} - \mu(z_1, z_2) )^2 \right] \partial z}{z_2 - z_1}
\end{equation}

%
We also define as \(\mathcal{Z}_s\) a sequence of \(K+1\) points that
partitions the \(s\)-th feature axis into a sequence \(K\)
variable-size intervals, i.e. \(\mathcal{Z}_s = \{z_0, z_K\}\), where
\(z_0=x_{s,min}\) and \(z_K=x_{s,max}\). Based on the above, we
redefine NAME using an interval-based formulation
\(\tilde{f}^{\mathtt{METHOD}}(x_s):=
\tilde{f}^{\mathtt{ALE}}_{\mu}(x_s) \pm \tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s)\), where:

\begin{equation}
  \label{eq:ALE_2}
  \tilde{f}^{\mathtt{ALE}}_{\mu}(x_s) = \sum_{k=1}^{k_x} \mu(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}

\begin{equation}
  \label{eq:ALE_accumulated_var}
  \tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s) =  \sum_{k=1}^{k_x} \sigma(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}
%
\paragraph{Theorem 1.}
\label{sec:theorem-1}

If we define the residual \(\rho(z)\) as the
difference between the expected effect at \(x_s\) and the mean
expected effect at the interval, i.e
\(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the accumulated variance
\(\sigma^2(z_1, z_2)\) equals to the accumulation of the point-wise
variances plus the squared residuals inside the interval:
\begin{equation}
 \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z
\end{equation}
%
The proof is in the Appendix.

\paragraph{Theorem 2.}
\label{sec:theorem-2}

The two definitions of NAME are equivalent, i.e.,
\(f^{\mathtt{ALE}}_{\mu}(x_s) = \tilde{f}^{\mathtt{ALE}}_{\mu}(x_s)\)
and
\(f^{\mathtt{ALE}}_{\sigma}(x_s) =
\tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s)\), if and only if the sequence
of limits \(\mathcal{Z}\) is such that
\(\sum_{k=1}^K \rho^2(z_{k-1}, z_{k}) = 0\), where \(\rho^2(z_{k-1}, z_{k}) = \int_{z_{k-1}}^{z_k} \rho^2(z)\partial z\).

\paragraph{Discussion}

Theorem 1 decouples the aggregated uncertainty inside a bin, i.e.,
\(\mathcal{H}_{bin}(z_1, z_2) := \sigma^2(z_1, z_2)\), into two
terms. The first term quantifies the accumulated uncertainty due to
the natural characteristics of the experiment, i.e.,
\(\mathcal{H}(z_1, z_2) = \int_{z_1}^{z_2}\mathcal{H}(z) \partial z\),
and the second term adds extra nuisance uncertainty due to limiting
the resolution
\(\mathcal{H}_n(z_1, z_2) := \int_{z_1}^{z_2}\rho^2(z) \partial z
\). Therefore, it holds that:
\begin{equation}
  \label{eq:uncertainty_bin}
  \mathcal{H}_{bin}(z_1, z_2) = \mathcal{H}(z_1, z_2) + \mathcal{H}_n(z_1, z_2)
\end{equation}
%
In Theorem 2 we prove that if we find a sequence of bins, where each
bin adds no nuisance uncertainty, then we will get exactly the same
NAME explanation as in the original definition.

@@

In this section we formulate bin-splitting as an unsupervised
clustering problem. The mean effect (Eq.~\eqref{eq:mu_bin}) and the
uncertainty (Eq.~\eqref{eq:var_bin}) can be directly estimated from
the set \(\mathcal{S}\) of the dataset instances with the \(s\)-th
feature lying inside the interval
\( \mathcal{S}= \{ \mathbf{x}^i : z_1 \leq x^i_s < z_2 \} \), i.e.
\(\hat{\mu}(z_1, z_2) = \frac{1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i
  \in \mathcal{S}} \left [ \dfdx(\mathbf{x}^i) \right ] \) and
\( \hat{\sigma}^2(z_1, z_2) = \frac{1}{|\mathcal{S}|}
\sum_{i:\mathbf{x}^i \in \mathcal{S}} \left ( \dfdx(\mathbf{x}^i) -
  \hat{\mu}(z_1, z_2) \right )^2 \). Both approximations are unbiased
only under the assumption that the points are uniformly distributed in
\([z_1, z_2]\). Under this assumption, and due to the law of large
numbers, the approximations become more robust, as the bins become
wider. On the other hand, bins should group local explanations with
similar effects and enlarging a bin is under the danger of adding
nuisance uncertainty. Therefore, we set-up the following optimization
problem:
\begin{equation}
  \label{eq:opt}
\begin{aligned}
  \min_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \quad & \mathcal{L} = \sum_{k=1}^K \hat{\sigma}^2(z_{k-1}, z_k) \Delta z_k \\
  \textrm{where} \quad & \Delta z_k = z_k - z_{k-1} \\
  & \tau_k = 1 - 0.2 \frac{|S_k|}{N} \\
  \textrm{s.t.} \quad & |\mathcal{S}_k| \geq N_{\mathtt{NPB}}\\
                                     & z_0 = x_{s,min}\\
                                     & z_K = x_{s, max}
\end{aligned}
\end{equation}
%
We search for the sequence of intervals \(z_0, \ldots, z_K\) that
minimizes the sum of the cost of each bin. The cost of each bin is the
mean variance \(\sigma^2_k\) scaled by the bin length \(\Delta z_k\)
and discounted by the term \(\tau_k\). The term
\(0.8 \leq \tau_K \leq 1\) acts as a tie-braker in cases of two
candidate bins with equal mean variance, favoring the selection of the
bigger bin. The constraint of at least \(N_{\mathtt{PPB}}\) points per
bin sets the lowest-limit for a \textit{robust} estimation.
